---
output: 
  github_document:
    html_preview: false
---

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/medrc)](https://cran.r-project.org/package=medrc)
[![Build Status](https://travis-ci.org/daniel-gerhard/medrc.svg?branch=master)](https://travis-ci.org/daniel-gerhard/medrc)
[![Downloads](https://cranlogs.r-pkg.org/badges/medrc)](https://cranlogs.r-pkg.org/)
 

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# medrc

## Overview

The medrc package is an extension to the package drc that allows you to fit hierarchical dose-response models and provides inferenc for derived parameters like the effective dose or benchmark doses.
Parameters are estimated with a two-stage approach using the package metafor, or using the Lindstrom-Bates algorithm implemented in the package nlme.

## Installation

```{r, eval = FALSE}
## You can install medrc from GitHub
# install.packages("devtools")
## first installing drc and drcData
devtools::install_github("daniel-gerhard/drc")
devtools::install_github("daniel-gerhard/drcData")
## then installing the development version of medrc
devtools::install_github("daniel-gerhard/medrc")
```
